---
title: "MiniProject Final"
author: "Isobel Bender"
date: "2024-02-20"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Set Up
```{r}
rm(list=ls())
#install.packages("ggbreak")
#install.packages("dplyr")
library(dplyr)
library(ggbreak) 
library(ggplot2)
library(car)
require(usdm)
require(psych)
require(lmerTest)
require(sjPlot)
```


#Data Loading and Cleaning
```{r}
#load data from the AVONET csv file 
AVONET_full <- read.csv("AVONET_full.csv")
#change HWI label for easier coding
#AVONET_full$Hand.Wing.Index <- AVONET_full$`Hand-Wing.Index`
#log transform Range size since it has a gamma distribution
AVONET_full$Log.Range <- log(AVONET_full$Range.Size)
#Subset to make df of only the Cracidae family


AVONET_Cracidae <- AVONET_full %>%
  filter(Family3=="Cracidae")


#remove island endemic species
AVONET_Cracidae <- AVONET_Cracidae[which(AVONET_Cracidae$Species3!="Pipile pipile"), ]
#remove species extinct in the wild
AVONET_Cracidae <- AVONET_Cracidae[which(AVONET_Cracidae$Species3!="Mitu mitu"), ]
```

Now we have a subsetted data frame with only the cracids of interest.

#Preliminary Data Visualization

##Histograms
This is the range size not log transformed- has similar to a gamma distribution (continuous, positive only).
```{r}
par(mfrow=c(1,2))
hist(AVONET_full$Range.Size, breaks=50)
hist(AVONET_Cracidae$Range.Size, breaks=8)
```
Here is the histograms of the log transformed range size - more normally distributed so better for linear modelling.
```{r}
par(mfrow=c(1,2))
hist(AVONET_full$Log.Range, breaks=50)
hist(AVONET_Cracidae$Log.Range, breaks=20)
```

## Histogram for report
```{r}
#filtering out the entries with NA for Range Size
AVONET_noNA <- AVONET_full %>%
  filter(Range.Size != "NA")

#Plot
ggplot(AVONET_noNA, aes(x=Range.Size)) + 
  geom_histogram(bins=150, fill="hotpink")+
  labs(title="Histogram of Bird Species Range Size (n=9855)",
       y="Number of Species")+ theme_bw()+
  scale_x_continuous(
    name="Range Size (km^2)",
    breaks=c(0,50000000,100000000),
    labels=c("0","50,000,000","100,000,000"))+
  theme(axis.text.x = element_text( 
                           size=10, angle=0))

```

##PCA 
PCA on all 11 morphological traits. 
```{r}
cracid_pca <- princomp(AVONET_Cracidae[ , c(9:19)],cor=TRUE)
summary(cracid_pca)
```
###Latent Root Criterion
```{r}
eigenvalues <- cracid_pca$sdev^2
eigenvalues
```
Comp1 and Comp2 are sufficient. Others would be dropped as they are below 1.  

###Scree Plot Criterion 
```{r}
plot(cracid_pca, type="lines", ylim=c(0,10))
```
Elbow to me seems to be at like 2

###Relative percent variance criterion 
```{r}
summary(cracid_pca)
```
We want at least 70. This criterion says only Comp1. But the other two tests say 2 comps, so 2 comps used going forward.

```{r}
loadings(cracid_pca)
```

Comp1 is approximately Size/Heftiness/Big beak - "Beak Size and Body Size"
Comp2 is Wing length/HWI/KD - Wing Shape and Dispersal Ability 
See AVONET, Tobias 2022 - "The Kipp distance is a metric used to measure wing elongation in birds, which is related to flight efficiency and dispersal ability."

```{r}
biplot(cracid_pca) 
```

### PCA Plots
```{r}
cracid_comps<- cbind(AVONET_Cracidae, cracid_pca$scores[,c(1,2)])

PCA_plot_range <- ggplot(data=cracid_comps, aes(x=Comp.1, y=Comp.2, 
                                                   col=Range.Size))+
  geom_point(size=4)+
  scale_color_viridis_b()+
  theme_bw()+
  labs(title="PCA of Cracids")+
  xlab("Component 1 (Beak Size and Body Size)")+
  ylab("Component 2 (Wing Shape and Dispersal Ability)")+
  geom_hline(yintercept=0, 
                color = "black", size=0.5)+
  geom_vline(xintercept=0, 
                color = "black", size=0.5)

print(PCA_plot_range) 
```

### PCA LM
```{r}

lm4<- lm(Log.Range~Comp.1*Comp.2, data = cracid_comps)
summary(lm4)

lm4.1<- lm(Log.Range~Comp.1+Comp.2, data = cracid_comps)
summary(lm4.1)

lm4.2<- lm(Log.Range~Comp.1, data = cracid_comps)
summary(lm4.2)

lm4.3<- lm(Log.Range~Comp.2, data = cracid_comps)
summary(lm4.3)

#although not significant, I want to plot them to see
par(mfrow=c(1,2))
plot(Log.Range~Comp.1,data = cracid_comps)
abline(lm4.2)
plot(Log.Range~Comp.2,data = cracid_comps)
abline(lm4.3)

```
###Inspect - Plots and Outliers
```{r}
cooksDPCA <- cooks.distance(lm4)
influential <- cooksDPCA[(cooksDPCA > (3 * mean(cooksDPCA, na.rm = TRUE)))]
influential
```
```{r}
par(mfrow=c(2,2))
plot(lm4)
```


###Remove Outliers
```{r}
#This corresponds to entry 14 and 44
cracid_comps[c(13,29,41),]

#removing the outliers (although I believe they are valid, I just want to see how much influence they are exerting on the model - is the model signficant without them?)
cracid_comps_no_out<- cracid_comps[-c(13,29,41),]

lm4.0.1<- lm(Log.Range~Comp.1*Comp.2, data = cracid_comps_no_out)
summary(lm4.0.1)

lm4.0.2<- lm(Log.Range~Comp.1, data = cracid_comps_no_out)
summary(lm4.0.2)

lm4.0.3<- lm(Log.Range~Comp.2, data = cracid_comps_no_out)
summary(lm4.0.3)

#None are significant, still
```

#Supplementary Work 
This is code I wanted to save and be able to reference in the future. None was included in the report.

##LM Just HWI

### in Cracids 
```{r}
#CRACIDS: Linear Model log(Range) vs. HWI
lm1 <- lm(Log.Range~Hand.Wing.Index, data=AVONET_Cracidae)
summary(lm1)
# diagnostic/residual plots  
par(mfrow=c(2,2))
plot(lm1)
# look OK - assumptions not violated - no Cook's distance outliers
```

### in all birds
```{r}
#HWI in 
lm1.1 <- lm(Log.Range~Hand.Wing.Index, data=AVONET_full)
summary(lm1.1)
#significant
#diagnostic plots
par(mfrow=c(2,2))
plot(lm1.1)
# no violations 

par(mfrow=c(1,1))
plot(Log.Range~Hand.Wing.Index, data=AVONET_full)
abline(lm1.1, col="red")
```
HWI does increase range size in all birds. R2 in all birds is 0.03. 

### all birds w/ small HWI
Subset all birds to just the birds with a relatively small HWI (similar to Cracids) - does the relationship still stand?
Yes it does.
```{r}
AVONET_small_HWI <-  AVONET_full %>%
  filter(Hand.Wing.Index<21)
  
lm1.2 <- lm(Log.Range~Hand.Wing.Index, data=AVONET_small_HWI)
summary(lm1.2)

plot(Log.Range~Hand.Wing.Index, data=AVONET_small_HWI, pch=19, cex=0.1)
abline(lm1.2, col="red")
```
## LM Just mass
Not significant
```{r}
lm1.4 <- lm(Log.Range~Mass, data=AVONET_Cracidae)
summary(lm1.4)

plot(lm1.4)
```
```{r}
lm1.5 <- lm(Log.Range~Mass, data=AVONET_full)
summary(lm1.5)

plot(lm1.5)

plot(Log.Range~Mass, data=AVONET_full)
abline(lm1.5)
```

##LM All cracids, all variables 

### VIF reduction
There are 11 morphological variables in AVONET - many of these are incredibly collinear, so they must be reduced. I chose the threshold of 3. 

```{r}
AVONET_Cracidae <- as.data.frame(AVONET_Cracidae)
vif(AVONET_Cracidae[9:19])
vif(AVONET_Cracidae[c(9:15,17,18,19)])
vif(AVONET_Cracidae[c(9:13,15,17,18,19)])
vif(AVONET_Cracidae[c(9,11:13,15,17,18,19)])
vif(AVONET_Cracidae[c(9,11:13,15,18,19)])
vif(AVONET_Cracidae[c(9,11:12,15,18,19)])
vif(AVONET_Cracidae[c(11:12,15,18,19)])
vif(AVONET_Cracidae[c(11,15,18,19)])
```
```{r}
lm2 <- lm(Log.Range ~ Beak.Width*Kipps.Distance*Tail.Length*Mass, data=AVONET_Cracidae)
summary(lm2)
#remove highest order interaction (4 way interaction)
lm2.1 <- lm(Log.Range ~ 
              Beak.Width*Kipps.Distance*Tail.Length
            +Beak.Width*Kipps.Distance*Mass
            +Beak.Width*Mass*Tail.Length
            +Mass*Kipps.Distance*Tail.Length
            +Beak.Width*Kipps.Distance
            +Beak.Width*Mass
            +Beak.Width*Tail.Length
            +Kipps.Distance*Tail.Length
            +Kipps.Distance*Mass
            +Tail.Length*Mass
            +Beak.Width
            +Kipps.Distance
            +Tail.Length
            +Mass, data=AVONET_Cracidae)
summary(lm2.1)
#remove BW:TL:M to get lm2.2
lm2.2 <- lm(Log.Range ~ 
              Beak.Width*Kipps.Distance*Tail.Length
            +Beak.Width*Kipps.Distance*Mass
            +Mass*Kipps.Distance*Tail.Length
            +Beak.Width*Kipps.Distance
            +Beak.Width*Mass
            +Beak.Width*Tail.Length
            +Kipps.Distance*Tail.Length
            +Kipps.Distance*Mass
            +Tail.Length*Mass
            +Beak.Width
            +Kipps.Distance
            +Tail.Length
            +Mass, data=AVONET_Cracidae)
summary(lm2.2)
#remove KD:TL:Mass to get lm2.3
lm2.3 <- lm(Log.Range ~ 
              Beak.Width*Kipps.Distance*Tail.Length
            +Beak.Width*Kipps.Distance*Mass
            +Beak.Width*Kipps.Distance
            +Beak.Width*Mass
            +Beak.Width*Tail.Length
            +Kipps.Distance*Tail.Length
            +Kipps.Distance*Mass
            +Tail.Length*Mass
            +Beak.Width
            +Kipps.Distance
            +Tail.Length
            +Mass, data=AVONET_Cracidae)
summary(lm2.3)
#We remove TL:Mass to get lm2.4
lm2.4 <- lm(Log.Range ~ 
              Beak.Width*Kipps.Distance*Tail.Length
            +Beak.Width*Kipps.Distance*Mass
            +Beak.Width*Kipps.Distance
            +Beak.Width*Mass
            +Beak.Width*Tail.Length
            +Kipps.Distance*Tail.Length
            +Kipps.Distance*Mass
            +Beak.Width
            +Kipps.Distance
            +Tail.Length
            +Mass, data=AVONET_Cracidae)
summary(lm2.4)
# all terms significant, nothing can be removed
```

###Inspect LM and outliers
```{r}
par(mfrow=c(1,1))
plot(lm2.4)
```

```{r}
cooksD <- cooks.distance(lm2.4)
influential <- cooksD[(cooksD > (3 * mean(cooksD, na.rm = TRUE)))]
influential
```

### LM without outliers 
The outliers were 6, 14, 30, 42. They were removed and the process restarted 
```{r}
#note - these don't match the numbers given above but for some reason they do select the outliers. 
AVONET_Cracidae[c(6,13,29,41),]

#removing the outliers 
AVONET_Cracidae_no_out <- AVONET_Cracidae[-c(6,13,29,41),]

lm2.4.1 <- lm(Log.Range ~ Beak.Width*Kipps.Distance*Tail.Length*Mass, data=AVONET_Cracidae_no_out)
summary(lm2.4.1)


lm2.4.2 <- lm(Log.Range ~ 
              Beak.Width*Kipps.Distance*Tail.Length
            +Beak.Width*Kipps.Distance*Mass
            +Beak.Width*Mass*Tail.Length
            +Mass*Kipps.Distance*Tail.Length
            +Beak.Width*Kipps.Distance
            +Beak.Width*Mass
            +Beak.Width*Tail.Length
            +Kipps.Distance*Tail.Length
            +Kipps.Distance*Mass
            +Tail.Length*Mass
            +Beak.Width
            +Kipps.Distance
            +Tail.Length
            +Mass, data=AVONET_Cracidae_no_out)
summary(lm2.4.2)
#remove BW:TL:M
lm2.4.3 <- lm(Log.Range ~ 
              Beak.Width*Kipps.Distance*Tail.Length
            +Beak.Width*Kipps.Distance*Mass
            +Mass*Kipps.Distance*Tail.Length
            +Beak.Width*Kipps.Distance
            +Beak.Width*Mass
            +Beak.Width*Tail.Length
            +Kipps.Distance*Tail.Length
            +Kipps.Distance*Mass
            +Tail.Length*Mass
            +Beak.Width
            +Kipps.Distance
            +Tail.Length
            +Mass, data=AVONET_Cracidae_no_out)
summary(lm2.4.3)
#remove KD:TL:M
lm2.4.4 <- lm(Log.Range ~ 
              Beak.Width*Kipps.Distance*Tail.Length
            +Beak.Width*Kipps.Distance*Mass
            +Beak.Width*Kipps.Distance
            +Beak.Width*Mass
            +Beak.Width*Tail.Length
            +Kipps.Distance*Tail.Length
            +Kipps.Distance*Mass
            +Tail.Length*Mass
            +Beak.Width
            +Kipps.Distance
            +Tail.Length
            +Mass, data=AVONET_Cracidae_no_out)
summary(lm2.4.4)
#remove BW:KD:Mass
lm2.4.5 <- lm(Log.Range ~ 
              Beak.Width*Kipps.Distance*Tail.Length
            +Beak.Width*Kipps.Distance
            +Beak.Width*Mass
            +Beak.Width*Tail.Length
            +Kipps.Distance*Tail.Length
            +Kipps.Distance*Mass
            +Tail.Length*Mass
            +Beak.Width
            +Kipps.Distance
            +Tail.Length
            +Mass, data=AVONET_Cracidae_no_out)
summary(lm2.4.5)
#remove BW:KD:TL
lm2.4.6 <- lm(Log.Range ~ 
              Beak.Width*Kipps.Distance
            +Beak.Width*Mass
            +Beak.Width*Tail.Length
            +Kipps.Distance*Tail.Length
            +Kipps.Distance*Mass
            +Tail.Length*Mass
            +Beak.Width
            +Kipps.Distance
            +Tail.Length
            +Mass, data=AVONET_Cracidae_no_out)
summary(lm2.4.6)
#remove BW:M
lm2.4.7 <- lm(Log.Range ~ 
            Beak.Width*Kipps.Distance
            +Beak.Width*Tail.Length
            +Kipps.Distance*Tail.Length
            +Kipps.Distance*Mass
            +Tail.Length*Mass
            +Beak.Width
            +Kipps.Distance
            +Tail.Length
            +Mass, data=AVONET_Cracidae_no_out)
summary(lm2.4.7)
#remove TL:M
lm2.4.8 <- lm(Log.Range ~ 
            Beak.Width*Kipps.Distance
            +Beak.Width*Tail.Length
            +Kipps.Distance*Tail.Length
            +Kipps.Distance*Mass
            +Beak.Width
            +Kipps.Distance
            +Tail.Length
            +Mass, data=AVONET_Cracidae_no_out)
summary(lm2.4.8)
#remove BW:KD
lm2.4.9 <- lm(Log.Range ~ 
            Beak.Width*Tail.Length
            +Kipps.Distance*Tail.Length
            +Kipps.Distance*Mass
            +Beak.Width
            +Kipps.Distance
            +Tail.Length
            +Mass, data=AVONET_Cracidae_no_out)
summary(lm2.4.9)
#remove KD:M
lm2.4.10 <- lm(Log.Range ~ 
            Beak.Width*Tail.Length
            +Kipps.Distance*Tail.Length
            +Beak.Width
            +Kipps.Distance
            +Tail.Length
            +Mass, data=AVONET_Cracidae_no_out)
summary(lm2.4.10)
#remove BW:TL
lm2.4.11 <- lm(Log.Range ~ 
            Kipps.Distance*Tail.Length
            +Beak.Width
            +Kipps.Distance
            +Tail.Length
            +Mass, data=AVONET_Cracidae_no_out)
summary(lm2.4.11)
#remove TL:KD
lm2.4.12 <- lm(Log.Range ~ 
            Beak.Width
            +Kipps.Distance
            +Tail.Length
            +Mass, data=AVONET_Cracidae_no_out)
summary(lm2.4.12)
#remove BW
lm2.4.13 <- lm(Log.Range ~ 
            Kipps.Distance
            +Tail.Length
            +Mass, data=AVONET_Cracidae_no_out)
summary(lm2.4.13)
#remove TL
lm2.4.14 <- lm(Log.Range ~ 
            Kipps.Distance
            +Mass, data=AVONET_Cracidae_no_out)
summary(lm2.4.14)
#remove mass
lm2.4.15 <- lm(Log.Range ~ 
            Kipps.Distance, data=AVONET_Cracidae_no_out)
summary(lm2.4.15)
#nothing is significant
```

### Simpler Linear models, all points
Including just the four vif-reduced variables and then cutting down. All cracid data included (including the outliers).
```{r}
lm3 <- lm(Log.Range ~Beak.Width
            +Kipps.Distance
            +Tail.Length
            +Mass, data=AVONET_Cracidae)
summary(lm3)
#remove TL
lm3.1 <- lm(Log.Range ~Beak.Width
            +Kipps.Distance
            +Mass, data=AVONET_Cracidae)
summary(lm3.1)
#remove KD
lm3.2 <- lm(Log.Range ~Beak.Width
            +Mass, data=AVONET_Cracidae)
summary(lm3.2)
#remove Mass
lm3.1 <- lm(Log.Range ~Beak.Width, data=AVONET_Cracidae)
summary(lm3.1)
#Beak Width is significantly impacting range size.

plot(lm3.1)
#Outliers are the same, they are the main issue.

plot(Log.Range ~Beak.Width, data=AVONET_Cracidae)
abline(lm3.1)
```

```{r}
cooksD1 <- cooks.distance(lm3.1)
influential1 <- cooksD1[(cooksD1 > (3 * mean(cooksD, na.rm = TRUE)))]
influential1
```
Now only two outliers, 14 and 42. 


## PCA GLM
Just checking that a gamma link function with a GLM gives the same result as log transforming the Range Size. It does. 
```{r}
glm1<- glm(Range.Size~Comp.1*Comp.2, data = cracid_comps, family='Gamma')
summary(glm1)

glm2<- glm(Range.Size~Comp.1+Comp.2, data = cracid_comps, family='Gamma')
summary(glm2)

plot(glm1)
```
## GLM with only beak width 
This is interesting as it shows that beak width isn't significant in a GLM.
```{r}
glm3<- glm(Range.Size~ Beak.Width, data=AVONET_Cracidae, family='Gamma')

summary(glm3)
plot(glm3)

cooksD3 <- cooks.distance(glm3)
influential <- cooksD3[(cooksD3 > (3 * mean(cooksD, na.rm = TRUE)))]
influential

#only two outliers here, 14 and 21 
```


SO far I have discovered that there is no relationship between traits and range size. Sad. 


## Full avonet PCA and LMs
```{r}
full_pca <- princomp(AVONET_full[ , c(9:19)],cor=TRUE)
summary(full_pca)
#two comps
eigenvalues <- full_pca$sdev^2
eigenvalues
#three comps
plot(full_pca, type="lines", ylim=c(0,10))
#two comps
```
Two comp sufficient

```{r}
biplot(full_pca) 
```

```{r}
full_comps<- cbind(AVONET_full, full_pca$scores[,c(1,2)])

lm_full<- lm(Log.Range~Comp.1*Comp.2, data = full_comps)
summary(lm_full)

lm_full1<- lm(Log.Range~Comp.1+Comp.2, data = full_comps)
summary(lm_full1)

lm_full2<- lm(Log.Range~Comp.1, data = full_comps)
summary(lm_full2)

lm_full3<- lm(Log.Range~Comp.2, data = full_comps)
summary(lm_full3)
```

```{r}
loadings(full_pca)
```
Comp1 is beak shape, wing length, secondary, overall beak shape
Comp2 is -HWI, -KD, Mass, Tarsus - heftiness/big and bad at dispersal

```{r}
par(mfrow=c(1,2))
plot(Range.Size~Comp.1, data = full_comps)
abline(lm_full2, col="red")
plot(Range.Size~Comp.2, data = full_comps)
abline(lm_full3, col="red")
```

##Latitude (Rapoport's Rule)

```{r}
vif(AVONET_Cracidae[c(31:33)])
vif(AVONET_Cracidae[c(31:32)])
```


```{r}
lm_minlat <- lm(Log.Range ~ Min.Latitude, data=AVONET_Cracidae)
summary(lm_minlat)

lm_maxlat <- lm(Log.Range ~ Max.Latitude, data=AVONET_Cracidae)
summary(lm_maxlat)

lm_lat <- lm(Log.Range ~ Min.Latitude+Max.Latitude, data=AVONET_Cracidae)
summary(lm_lat)

par(mfrow=c(1,2))
plot(Log.Range ~ Min.Latitude, data=AVONET_Cracidae)
abline(lm_minlat)

plot(Log.Range ~ Max.Latitude, data=AVONET_Cracidae)
abline(lm_maxlat)
```


```{r}
vif(AVONET_full[c(31:33)])
vif(AVONET_full[c(31:32)])
```

```{r}
lm_minlatfull <- lm(Log.Range ~ Min.Latitude, data=AVONET_full)
summary(lm_minlatfull)

lm_maxlatfull <- lm(Log.Range ~ Max.Latitude, data=AVONET_full)
summary(lm_maxlatfull)

par(mfrow=c(1,2))
plot(Log.Range ~ Min.Latitude, data=AVONET_full)
abline(lm_minlatfull,col="red")

plot(Log.Range ~ Max.Latitude, data=AVONET_full)
abline(lm_maxlatfull,col="red")
```
```{r}
R.Version()
```

